Powering Hidden Markov Model by Neural Network based Generative Models

### 2019-08-28 manuscript/main.tex
update

HMM provides a framework for sequential data modeling. To connect states of HMM with input data, a probabilistic model is used to describe how each state of HMM should represent a pattern of sequential data. The typical way is to use a Gaussian mixture model (GMM) per state of HMM \cite{juang1986maximum}, where GMMs are used to represent the relationship between states of HMM and sequential data input. GMM based HMM (GMM-HMM) has becomes a standard model for sequential data modeling, and employed widely for practical applications, especially in speech recognition \cite{gales2008application,chatterjee2011auditory}.

@@

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybrid. A hierarchical training is carried out by: i) training a HMM first, ii) and then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn.

@@

In this work, we propose a generative model based HMM. Specifically, we propose a generator-mixed HMM, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
  \item firstly has high modeling capacity of sequential data, due to the neural network based generators;
\item has advantage of easy to do training. Training of GenHMM employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible;
% \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

@@

Our framework is a Hidden Markov Model (HMM). A HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for difference sequential data can be different.}. Using HMM for signal representation is illustrated in Figure~\ref{fig:hmm}. The model assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by a different signal source associated with a hidden state of HMM, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of hidden states of HMM $\bm{H}$.
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is initial distribution of HMM $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For a state $s$, $q_s = p(s;\bm{H})$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$ and $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For given hidden state $s$, the density function of observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model. Denote $\bm{\Phi} = \left\{ \bm{\Phi}_{s}| s \in \Ss \right\}$.
\end{itemize}

>>>>

HMM provides a framework for sequential data modeling. To connect states of HMM with input data, a probabilistic model is used to describe how each state of HMM represents a pattern of sequential data. The typical way is to use a Gaussian mixture model (GMM) per state of HMM \cite{juang1986maximum}, where GMMs are used to represent the relationship between states of HMM and sequential data input. GMM based HMM (GMM-HMM) has becomes a standard model for sequential data modeling, and employed widely for practical applications, especially in speech recognition \cite{gales2008application,chatterjee2011auditory}.

@@

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybrid. A hierarchical training is carried out by: i) training a HMM first, ii) then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn.

@@

In this work, we propose a generative model based HMM. Specifically, we propose a generator-mixed HMM, termed as GenHMM, where generators are realized by neural networks to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
  \item has high modeling capacity of sequential data, due to the neural network based generators;
\item has advantage of easy to do training. Training of GenHMM employs expectation maximization (EM) framework. Training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible;
% \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are implemented as flow models.

@@

Our framework is a HMM. A HMM $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for difference sequential data can be different.}. Using HMM for signal representation is illustrated in Figure~\ref{fig:hmm}. The model assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by a different signal source associated with a hidden state of HMM, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of hidden states of HMM $\bm{H}$.
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is the initial distribution of HMM $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For a state $s$, $q_s = p(s;\bm{H})$. In the following, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$ and $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For a given hidden state $s$, the density function of the observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model. Denote $\bm{\Phi} = \left\{ \bm{\Phi}_{s}| s \in \Ss \right\}$.
\end{itemize}
